---
title: "Factor codings for models in R"
author: "Heidi Seibold"
layout: post
category: Rbloggers
tags: [R, lm, glm, contrasts]
comments: True
---

I am holding an exercise on generalised models these days. Preparing a task on factor coding
in generalised linear models, I realised that the help on the internet on that is not so
easy to understand. At least what I found. So in order to help people who find this topic 
confusing, I want to help out a little here.

Consider the following data

```{r, fig.height=5, message=FALSE}

set.seed(29)
x <- gl(5, 10) # 5 factor's levels, each replicated 10 times
y <- rnorm(n = length(x), mean = as.integer(x), sd = 0.1)   # means= 1,2,3,4,5
dd <- data.frame(x = x, y = y)

library(ggplot2)
ggplot(aes(x = x, y = y), data = dd) + geom_boxplot()

means <- tapply(y, x, mean)
means

```


Now we want to model the mean of y given x using the `lm()` function with the following codings:
*dummy-coding*, *treatment-coding* (where the reference category is 5), *effect-coding*
and *split-coding*.

To make the theory more general, we have a categorical variable  \\( X \\) with \\( K \\) categories \\( (a\_1, \dots, a\_K) \\)


## Dummy-coding
Look at each level separately:
\\[ E(Y|X=a\_{k}) = \beta\_k, \quad k=1,\dots,K \\]

```{r}
# dummy (each level singularly)
lm_dummy <- lm( y ~ x - 1, contrasts = list(x = contr.treatment(5)))
summary(lm_dummy)

coef(lm_dummy)
means
```

## Treatment-coding
Compare each category to the dummy-category \\( a\_d \\):

\\[ E(Y|X=a\_{d}) = \beta\_0 \\]

\\[ E(Y|X=a\_k) = \beta\_0 + \beta\_k, \quad k\neq d \\]
```{r}
# treatment (restrict one level to constant term, all other difference from it)
lm_treatment <- lm( y ~ x, contrast = list(x = contr.treatment(5, base = 5)))
summary(lm_treatment)

coef(lm_treatment)
c(means[5], means[1:4] - means[5]) 
```


## Effect-coding
Compare each category to the mean:

\\[ E(Y|X=a\_k) = \beta\_0 + \beta\_k, \quad k=1,\dots,K-1 \\]

\\[ E(Y|X=a\_K) = \beta\_0 - \sum\limits\_{j=1}^{K-1} \beta\_j \\]
```{r}
# effect(deviation from overall average)
lm_effect <- lm(y ~ x, contrast = list(x = contr.sum(5)))
summary(lm_effect)

coef(lm_effect)
c(mean(means), means[1:4] - mean(means))
```


## Split-coding
Compare each category to the previous category (for ordered categories):

\\[ E(Y|X=a\_1) = \beta\_0 \\]

\\[ E(Y|X=a\_k) = \beta\_0 + \sum\limits\_{j=1}^{k-1} \beta\_j, \quad k=2,\dots,K \\]
```{r}
# split coding
c <- rbind( c(0,0,0,0),
            c(1,0,0,0),
            c(1,1,0,0),
            c(1,1,1,0),
            c(1,1,1,1) )

lm_split <- lm(y ~ x, contrast = list(x = c))
summary(lm_split)


coef(lm_split)
c(means[1], diff(means))
```








